---
title: "Final Project"
author: "Paul Markley"
date: "2022-11-29"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble
This script will take in any dataframe with a column 'species' that also has the columns latitude, longitude, year, month, and date. It then cleans the species identity based on the World Flora Online (WFO) taxonomic backbone. After, the script plots the counts of species and genus in the data frame, followed by spatial plotting. It lastly runs a mantel randomization test on the date and locations to see if there is any correlation between the space and time dimensions of the data. 

This script is intended to be used with a dataset downloaded from GBIF.

Below is a list of packages needed.

```{r}

if (!require("data.table", quietly = TRUE))
  install.packages("data.table")
if (!require('tidyverse', quietly = T))
  install.packages('tidyverse')
if (!require('ade4', quietly = T))
  install.packages('ade4')
if (!require('sf', quietly = T))
  install.packages('sf')
if (!require('terra', quietly = T))
  install.packages('terra')
if (!require('WorldFlora', quietly = T))
  install.packages('WorldFlora')

```

# Data Step
Now the data will be read in. These files probably are not in the repository since they are too big for GitHub. These will be called from the local location after setting the working directory.

```{r}
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
df <- read.csv('/Users/pm/COURSES/COMPBIO/grad-project/data/heaths.csv') # On onedrive
back <- data.table::fread("/Users/pm/COURSES/COMPBIO/grad-project/data/classification.txt") # on onedrive

```

# Taxonomic Resolution
After reading in the data I create the intersection of the species vectors. It is the list of species IDs that are OK and don't need correction. Using the function, I pick the rows from the clean.sp function that don't match any of the values in the intersected vector and call the unique function on that vector. Then the clean.sp function uses WFO.match's fuzzy matching algorithms to match species that may have been misspelled or ones with outdated names with the correct name. This vector is then matched back to the extracted rows and integrated back into the input data frame, which is returned at the end of this function.

```{r}

# remove rows that have matched with backbone, then gets all unique species names
# takes these species names and matches with accepted taxonomy.
S <- intersect(df$species, back$scientificName)
source('/Users/pm/COURSES/COMPBIO/grad-project/scripts/cleansp.R')
df <- clean.sp(df, sp.list = S)

# Final check: if it worked the number of species matched to the backbone should
# equal to the number of species in the dataset and prints TRUE.
S <- intersect(df$species, back$scientificName); length(S) == length(unique(df$species))

```

# Date of Year Calculation
I calculate the date of year (DOY) by calling lubridate's yday after collapsing the year, month, and day columns in the data frame into one column and updating the vector's class to Date. This will be used later in the script and for plotting the temporal trends in the data.

```{r}
# Convert the YMD columns to DOY for plotting and brief analysis.
DOY = lubridate::yday(as.Date(paste0(df$year, '-', df$month, '-', df$day)))
df$DOY = DOY
```

# Plotting

I want to look at the species and genus observation totals. So I call my second function in the freqsptbl.R file. It's a simple function that calls the table function on the string vector and then creates a genus column from the species binomial column. It also includes a function to filter the table by number of observations.

```{r}
# Making a few plots to visualize what happened here in the data.
library(tidyverse)

# This script contains a function that summarizes the data frame into a table by species
# It then makes a genus column from the species binomial.
source('/Users/pm/COURSES/COMPBIO/grad-project/scripts/freqsptbl.R')
tbl <- freq.sp.tbl(df)

# Now the plots
# First is frequency of species colored by genus
 ggplot(tbl, aes(y = Var1, x = Freq)) +
   geom_col(aes(fill = genus)) +
   theme_bw() +
   labs(y='Species', x= 'Number of Observations', title = 'GBIF observations by Species')
#ggsave('../figures/Speciestbl.png')
```

```{r}
#Second is just genus frequencies
ggplot(tbl, aes(y = genus, x = Freq)) +
 geom_col(aes(fill = genus)) +
 theme_bw() +
 labs(y='Genus',x = 'Number of Observations', title = 'GBIF observations by Genus') 
#ggsave('../figures/genustbl.png')
```

```{r}
# Third is a density plot of DOY by genus
ggplot(data = df, aes(x = DOY, fill = genus)) +
geom_density(alpha = 0.3) + 
theme_bw() +
labs(title = 'Density Plot of Date of Year by Genus')
#ggsave('../figures/densdoy.png')

#Fourth is a density plot of year by genus.
ggplot(data = df, aes(year, color = genus)) +
   geom_density() +
   labs(title = 'Density Plot of Year by Genus') +
  theme_bw()
#ggsave('../figures/densyear.png')

```

## Spatial Plots
Here I visualize the spatial patterns in the data to answer the question of: where are the most observations in the dataset?

```{r}
# Making GIS maps. This requires the sf and terra packages.
# I make from the dataframe a sf object in WGS84
sf1 <- sf::st_as_sf(df, coords = c('lon','lat'), crs = 'EPSG:4326')
ggplot(data = sf1)+
geom_sf()
#ggsave('../figures/map.png')
```

The figure that I made with geom_sf() is great, but it's hard to read. I have here next a rasterization of the data points. What is happening here is that I am binning number of points in each grid cell to produce a value to be plotted onto the map. The distribution of this new data set can be considered akin to a zero inflated Poisson, so I logged the values to make them a bit more readable and normally distributed. 

```{r}                                        
#Next I make a blank raster to rasterize the number of observations within each cell.
#The resolution of each cell is 50km by 50km, roughly a degree by a degree.
rst <- terra::rast(nrows = 57, ncols = 720, nlyrs = 1, resolution = c(0.5,0.5),
                   xmin = -180, xmax = 180, ymin = 55.5, ymax = 84, crs = 'EPSG:4326')

# I take the log of the observations to make the plot a bit easier to read.
abd <- log(terra::rasterize(terra::vect(sf1), rst, fun=length, touches = T))
plot(abd)
#png(filename = '../figures/rasterlogobs.tif', width = 500, height = 275)
#plot(abd)
#dev.off()
```

# Analysis
For the analysis I wanted to examine the relationship between location and time, or the correlation between the spatial and temporal dimensions. I did this first by making a distance matrix and then computing the Mantel test statistic from the ade4 package.

```{r}
# My dataset is too big, so I will reduce it with a predetermined subset of observations
# Original dataset: ~20,000 observations, restricted data has ~4,000
# These were based on whether or not the points were within a polygon.
df1 <- df[df$in_s %in% 'PM',]
df1 = df1[complete.cases(df1) == T,]

# Minor Analysis of DOY vs Location and Year vs Location using the Mantel randomization test
d1 <- dist(cbind(df1$lon, df1$lat))
d2 <- dist(df1$DOY)
d3 <- dist(df1$year)

(mtl <- ade4::mantel.rtest(d1, d2))
saveRDS(mtl, '/Users/pm/COURSES/COMPBIO/grad-project/figures/mantel_out.RData')
# H0: Location and DOY are not linearly correlated.
# Ha: Location and DOY have some linear correlation.
# pvalue = 0.01 < alpha 0.05: reject null hypothesis
# There is sufficient evidence to conclude that location and DOY are significantly positively correlated.

(mtl2 <- ade4::mantel.rtest(d1,d3))
saveRDS(mtl2, '/Users/pm/COURSES/COMPBIO/grad-project/figures/mantel_out2.RData')
# H0: Location and year are not linearly correlated.
# Ha: Location and year have some linear correlation.
# pvalue = 0.01 < alpha 0.05: reject null hypothesis.
# There is sufficient evidence to conclude that location and year are significantly positively correlated.
```